---
title: "PSY 613, Spring 2021: Lab 2 - Programming in R"
date: "4/8/2022"
output:
  html_document: 
    fig_caption: yes
    theme: cosmo
    toc: yes
    toc_depth: 3
    toc_float: TRUE
    df_print: paged
editor_options: 
  chunk_output_type: console
---
  
---

  
```{r setup, include=FALSE}
# I always include a setup chunk in my Markdown files: It helps me ensure that I've got my settings and packages straightened out before my script starts.

# Set chunk options
knitr::opts_chunk$set(echo = TRUE)

# If you haven't installed the packages, do that first by uncommenting the next line!
# install.packages(c("pracma", "knitr"))

# Then load the required libraries
library(pracma)
library(knitr)
```

_I like to include shortcuts I'm likely to forget up here..._  
 - Run the current line of code: `Cmd`+`Enter` (on a Mac) or `Ctrl`+`Enter` (on a PC).  
 - Run the current chunk: `Cmd`+`Shift`+`Enter` (on a Mac) or `Ctrl`+`Shift`+`Enter` (on a PC).  
 - Run all previous chunks: `Cmd`+`Opt`+`P` (on a Mac) or `Ctrl`+`Alt`+`P` (on a PC).  
 - Clear console: `Ctrl`+`L`. (Note: This command doesn't clear your workspace.)  
 

# Review   

### _What is a sampling distribution?_   
_Answer:_  
  


### _What is the central limit theorem (CLT)?_   
_Answer:_  
  
  
---
  
# Creating a Sampling Distribution  

_Where to begin...?_  

Before we compare some sampling distributions to demonstrate CLT and all that jazz, let’s write a script for a sampling distribution of the mean that consists of 10 samples with a sample size of 5. This will be our template script for the following chunks.   

```{r sampdist_prac}
# How do we indicate the number of iterations (i.e., the number of samples)?
# Save it as a variable!
num_samples <- 10

# How do we indicate the sample size?
# Save it as a variable!
sample_size <- 5

# For each of the 10 iterations, there will be a sample mean.  We need to create a vector that stores all of these sample means.  How do we do this?

# There are lots of different right answers to this. One easy way is to make a vector of the right size filled with NAs, and then we'll over-write each of the zeros as we get our sample means.

sample_means <- rep(NA_real_, 10) # Hint: Look up the function rep()
# 10 because we're taking 10 samples
  
# Check out what you just made. 
sample_means # It's a vector of 10 NAs, yay

# It should all be NAs for now

sum(is.na(sample_means)) # What's 10 NA's summed? 10!

## Question: Why would it be useful to use NAs when creating the storage vector?
## Answer: 


# -------------------------------------------------------------------------------------------------

# How do we tell R to run i, ii, and iii for each of the 10 iterations (this is known as a for-loop)?

## Question: What is 'i' in the for-loop?
## Answer: 

for (i in 1:num_samples) { #i in 1:10 to run through 10 times
  

  # How do we tell R to randomly select a sample with a sample size of 5 from a uniform distribution?
  sample <- runif(sample_size) # Get sample
  
  sample_mean <- mean(sample) #get mean of that sample 
  # Note that this variable is different from the storage vector we made above (called "sample_means")
  
  # Store the sample mean
  sample_means[i] <- sample_mean # Hint: you will need to index using i
  
}


# Now check out your sample_means variable
sample_means

# Can also check out the mean of our sample means!
mean(sample_means)

```
  
The mean of our sample means is `r round(mean(sample_means),2)`.  
  
---
  
# Demonstrating the CLT in R   


Let’s compare two sampling distributions of means (`sampdistA` & `sampdistB`), both of which consist of 100 samples from a uniform distribution.  Whereas `sampdistA` consists of samples with a sample size of 5, `sampdistB` consists of samples with a sample size of 50.   
    
### _Make a simulation with small samples (*n* = 5)_ 

```{r sampdistA}
num_samplesA  <- 1000 # Note that we're running more iterations now.  
sample_sizeA  <- 5
sample_meansA <- rep(NA, num_samplesA) #note the preallocation

for (i in 1:num_samplesA) {
  sampleA          <- runif(sample_sizeA)
	sample_meanA     <- mean(sampleA)
	sample_meansA[i] <- sample_meanA
}

hist(sample_meansA, 
     main = "Sampling Distribution with Sample Size 5", 
     xlab = "Sample Means")

# How would I change this histogram code to make the title specify the sample size based on a variable?

hist(sample_meansA, 
     main = paste("Sampling Distribution with Sample Size", sample_sizeA)) # Hint: Look up the function paste()

```
  
_Question:_ What part of this code would you need to change to make a simulation with a larger sample size?  
_Answer:_  
  
### _Make a new simulation with larger samples (*n* = 50)_   

```{r sampdistB}
# Note that when we run this code, we would overwrite all the variables we made before if we just used the same names. I changed "A" to "B" for the variables since I didn't want to overwrite the original ones. 
num_samplesB  <- 1000
sample_sizeB  <- 50 # This is the only real change in the code
sample_meansB <- rep(NA, num_samplesB)

for (i in 1:num_samplesB) {
  sampleB          <- runif(sample_sizeB)
  sample_meanB     <- mean(sampleB)
	sample_meansB[i] <- sample_meanB
}

hist(sample_meansB, main = paste("Sampling Distribution with Sample Size", sample_sizeB), xlab = "Sample Means")

```
  
_Question:_ Do the histograms of the two sampling distributions differ?  
_Answer:_  They're supposed to, one thing you can look out for is what looks like a triangular shape vs. a smooth histogram shape. The larger sample size is meant to be closer to a normal distribution so one thing you could look for is kurtosis.
  

### Now Let's Plot Some Stuff!  
  
_What if we wanted to display both histograms next to each other in the same figure?_   
  
Use the `par()` function. You run this before you generate plots to set the parameters for those figures. To tell it to include more than one plot in the same figure, use `par(mfrow)` to set up a matrix of plots. Set `mfrow = c(row,col)` where _row_ is the number of rows of plots you want, and _col_ is the number of columns. For example, `par(mfrow = c(2,3))` will set up a matrix of 6 plots, in 2 rows of 3 plots each.   

See [Quick R](http://www.statmethods.net/advgraphs/layout.html) for details and more examples.   
```{r hists}
par(mfrow = c(1,2))

# Our first histogram
hist(sample_meansA, 
     main = paste("N =", sample_sizeA), 
     xlab = "Sample Means")   

# Our second histogram
hist(sample_meansB, 
     main = paste("N =", sample_sizeB), 
     xlab = "Sample Means") 

```
  
_What if you wanted to overlay a normal curve on these, to see how well the histograms fit a normal distribution?_   

```{r hists_with_curve}
# Use curve(), and the function dnorm(), which uses the formula for the normal distribution. Remember the normal dist is defined by a mean and SD, so we have to include those. Use the means and SDs from your sampling distributions. 
meanA <- mean(sample_meansA)
sdA   <- sd(sample_meansA)
meanB <- mean(sample_meansB)
sdB   <- sd(sample_meansB)

par(mfrow = c(1, 2)) # Set figure to be 1 row, 2 columns

# Plot histogram of sample size 5
hist(sample_meansA, 
     freq   = FALSE, 
     breaks = 20, 
     main   = paste("N =", sample_sizeA), 
     xlab   = "Sample Means", 
     xlim   = 0:1)
# Add curve to the histogram
curve(dnorm(x, mean = meanA, sd = sdA), # you don't have to create x object; it knows
      col = "cornflower blue", 
      add = TRUE) 

# Plot histogram of sample size 50
hist(sample_meansB, 
     freq   = FALSE, 
     breaks = 20, 
     main   = paste("N =", sample_sizeB), 
     xlab   = "Sample Means", 
     xlim   = 0:1)
# Add curve to the histogram
curve(dnorm(x, mean = meanB, sd = sdB), 
      col = "cornflower blue", 
      add = TRUE)

# -------------------------------------------------------------------------------------------------

# What if I just wanted to plot one of the above graphs now?
hist(sample_meansB, 
     freq   = FALSE, 
     breaks = 20, 
     main   = paste("N =", sample_sizeB), 
     xlab   = "Sample Means",
     xlim   = 0:1) 

# Need to revert the par command:
par(mfrow = c(1,1))
# Or
dev.off()

# Plot
hist(sample_meansB, 
     freq   = FALSE, 
     breaks = 20, 
     main   = paste("N =", sample_sizeB), 
     xlab   = "Sample Means",
     xlim   = 0:1) 

```
  
_What if I wanted to just report mean of the sample means and the sample size (i.e., size of each sample that we used to create our distribution) and print it to the workspace?_

```{r means_on_means_on_means}
# we have this info from above:
num_samplesA # Should be 1000
sample_sizeA # Should be 5
sample_meansA # Is the vector of sample means we created using our 'for' loop

# We can calculate the mean of the sample means
msm <- mean(sample_meansA)

# Maybe I want to round it
msm_rounded <- round(msm, 2)

paste("With sample size", sample_sizeA, "the mean of the sample means is", msm_rounded, ".")
# note stupid space between sentence and your period.
# can try using paste0 instead with manual spaces
paste0("With sample size ", sample_sizeA, "the mean of the sample means is ", msm_rounded, ".")
```   

_Sidenote: If you prefer to write these types of statements outside of chunks in your Markdown files, you can do it like this:_    
With sample size `r sample_sizeA`, the mean of the sample means is `r msm_rounded`.  
  
    
---  
_What if I wanted to systematically compare lots of possible sample sizes?_

# Creating Multiple Sampling Distributions within a Nested For-Loop  

```{r multiple_sampledists}
# Nested for-loop!

# Set number of samples (iterations)
num_samples <- 1000 

# Set different sample sizes you want to compare
sample_sizes = c(5, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100, 110)

# Set up a matrix to plot the histograms
par(mfrow = c(3,4)) # Matrix with 3 rows and 4 columns

# We're going to pull over 12 different sample sizes and create a histogram with a label as well that specifies the sample size
for (i in 1:sample_sizes) { #looping over each sample size
  # Create vector to save samples
  sample_means <- rep(NA, num_samples) 
  
  for (j in 1:num_samples) { #take 1000 different samples of each different sample size
    
    sample          <- runif(sample_sizes[i]) # Get sample of specified sample size - note indexed i
	  sample_mean     <- mean(sample)
	  sample_means[j] <- sample_mean # Assign sample mean to correct place in vector

  }
  
  # Add the histogram
  hist(sample_means, 
       main = paste("Sampling Distribution with Sample Size", sample_sizes[i]), 
       xlab = "Sample Means")
  
  # If you wanted to print the mean of the sample means
  print(paste("Sample mean with N =", sample_sizes[i], "is", round(mean(sample_means), digits = 2)))

}
  
```


