---
title: 'Lab 9: Multi-Level Modeling'
date: "May 28, 2021"
output: 
  html_document:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(scipen = 99)

# install.packages("lme4")
library(rio)
library(tidyverse)
library(lme4)
library(here)
```

Today we'll be reviewing multi-level modeling and covering how to run MLMs in R. Across the examples today we'll be working with the NELS-88 dataset that Elliot worked with in class. Let's start by reading in that dataset.

# Data & Libraries
## Importing & Cleaning Data

We'll import the data, check its structure, and then do any necessary cleaning.
```{r get_data}

nels_88 <- import(here("data", "613_Lab09_NELS88.sav")) %>% 
  rio::factorize()

colnames(nels_88) <- tolower(colnames(nels_88))

head(nels_88)
str(nels_88)

```

Looks like our two ID variables (`schoolid` & `studentid`) are being read in as a numeric variable instead of a factor. We'll change those now to avoid any errors down the road.

```{r factorize_vars}
nels_88 <- nels_88 %>% 
  mutate(schoolid = factor(schoolid),
         studentid = factor(studentid))
```


# Precursors to MLM

Multi-level data are data which are clustered in some way. Some examples might include:

* Students in schools
     * Schools in (school) districts, or school districts in states
* Observations nested in an individual (i.e., repeated measures) 
* Individuals nested in groups in an experiment
* Individuals nested in dyads

We usually refer to the lower level units (e.g., students) as *Level 1* and the higher level units (e.g., schools) as *Level 2*. Before getting into MLM proper, we'll start by going through some of the precursors.

## Disaggregation vs. Aggregation

One approach to multi-level data is to just work with the data *only* at Level 1 or *only* Level 2, called disaggregation and aggregation, respectively.

Disaggregation                    | Aggregation
----------------------------------|--------------------------------
 Ignores Group-Level Data         | Ignores Individual-Level Data
 (discards b/w group variability) | (discards w/i group variability)
 

### Disaggregation

Estimate a disaggregated model predicting math achievement from time spent on math homework each week. Since disaggregation ignores group-level data, we will not use the schoolid variable in our regression model.

```{r disag_model1}
disag_model_1 <- lm(mathscore ~ timeonmath, data = nels_88)
summary(disag_model_1)
```

>Q: What do the results suggest?    

A: mathscores = 44.07 + 3.57*timeonmath

>Q: What is the problem with conducting the analysis in this manner?

A: Ignoring a grouping variable could inflate our degrees of freedom, which will result in inflating our chances of making a Type I error 


### Aggregation

Estimate an aggregated model predicting mean math achievement from mean hours spent on homework each week. Since aggregation ignores individual-level data, we need to compute the mean math score and the mean time spent on math homework for each of the 10 schools. We will do this in the `tidyverse` by using `group_by()` and `summarize()` from the `dplyr` library.

```{r aggregate}
nels_88_agg <- nels_88 %>% 
  group_by(schoolid) %>% 
  summarize(m_mathscore = mean(mathscore, na.rm = TRUE),
            m_timeonmath = mean(timeonmath, na.rm = TRUE))
```

And then we just run the regression with these new variables in our new dataset.

```{r agg_model1}
agg_model_1 <- lm(m_mathscore ~ m_timeonmath, data = nels_88_agg)
summary(agg_model_1)
```

>Q: What do the results suggest?     

A: mean_mathscores = 40.02 + 4.98*mean_timeonmath  

>Q: What is the problem with conducting the analysis in this manner?    

A: Decreased power, which increases chances of Type II error    


## Examining Variability in Intercepts and Slopes

It can be useful to visualize our data to see if they vary in slopes, intercepts, or both. We have a few options for visualizing this in ggplot, and your choice might depend on things like how many groups you have.

First, we can use `facet_wrap()` to generate a separate plot for each school:

```{r plot_all_facets}
ggplot(data = nels_88, aes(x = timeonmath, y = mathscore)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) + # tell ggplot to conduct a lm
  facet_wrap(~ schoolid) + # don't forget the ~
  theme_minimal() # I like this theme; totally optional
```

A second option is to plot them all on a single plot, and color code lines and points by school. In some ways, this is even easier; we can do this by using the `color` aesthetic.

```{r plot_all_together}
ggplot(data = nels_88, aes(x = timeonmath, y = mathscore, color = schoolid)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) + # tell ggplot to conduct a lm
  theme_minimal()
```

>Q: Does it look like there might be between-school variability in the intercepts, slopes (for time spent on math), or both?   

A: Both

## (Two-Part) Slopes & Intercepts as Outcomes

One option that Elliot mentioned in class yesterday is to basically run a two-stage analysis. First, we assess the extent to which intercepts and slopes vary (between schools in this case), and then we save those parameters and use them as outcomes in the next part 

### Part 1: Assessing Variability & Running OLS in each group

#### Step 1: Dummy Code (if necessary)

Since we will now include a categorical variable (to mark the groups) in our regression model, we need to create either dummy variables or make sure it is a factor in R and allow R to do the dummy coding. We already made sure school id was a factor above, and we don't have any particular school that should be the reference group, so we'll let R do our dummy coding for us. If we wanted to use some other coding scheme, e.g., `contr.sum(10)`, which implements effects coding, we would just assign that matrix to `contrasts(df$var)` 

*Note: you should always think about whether and how to code a categorical variable when including them in a regression.*


#### Step 2: Assess Between Group Variability in Intercepts

Next we'll check if schools differ in their intercepts by including `schoolid` in the model and see if it improves the model's fit.

In other words, we'll compare the following models:

Model 1: $MathAch = b_0 + b_1TimeOnMath$  
Model 2: $MathAch = b_0 + b_1TimeOnMath + b_2School_1 + ... + b_{10}School_9$  


```{r disag_model2}
disag_model_2 <- lm(mathscore ~ timeonmath + schoolid, data = nels_88)
anova(disag_model_1,
      disag_model_2)
```

>Q:  Looking at the tables below, do we have a main effect of our grouping variable (i.e., school)?  In other words, is the change in R squared significant? 

A: We do have a significant effect of school, F(9, 249) = 13.94, p < .001.   

>Q: What is another name for this model?    

A: ANCOVA    

#### Step 3: Assess Variability in Slope 

Next we'll want to determine if there is variability school-to-school in the effect of time spent on math homework on math achievement scores. We'll do this by including the interaction between `schoolid` and `timeonmath`. We'll be comparing the following models:

Model 2: $MathAch = b_0 + b_1TimeOnMath + b_2School_1 + ... + b_{10}School_9$  
Model 3: $MathAch = b_0 + b_1TimeOnMath + b_2School_1 + ... + b_{10}School_9 + b_{11}(School_1*TimeOnMath) + ... + b_{19}(School_9*TimeOnMath)$  


```{r disag_model3}
disag_model_3 <- lm(mathscore ~ timeonmath*schoolid,
                    data = nels_88)

anova(disag_model_2,
      disag_model_3)
```

Now we know that there is significant between-school variance in both the intercepts and the slopes, and we can try to predict this variance. To do this, we need to find the unique intercept and slope values for each of the 10 schools.

#### Step 4: Get an Intercept & Slope for each School

##### Getting intercepts and slopes by hand

We can calculate the slope and intercept for each school by hand from this model's coefficients.

```{r}
summary(disag_model_3)
levels(nels_88$schoolid)
```

Intercept: b0 = 50.68, predicted math score when timeonmath = 0 and when schoolid = 0 across all our dummy codes (when in the reference school)

timeonmath: b1 = -3.55, relationship between timeonmath and math scores when all our schoolid dummy codes = 0 (when in the reference school)

>Q: What is the unique intercept for school 7930?   

A: Intercept for school 7930 = 38.75
```{r}
50.68-11.93
```

    

>Q: What is the unique slope for school 7930?

A: Slope for school 7930 = 7.91
```{r}
-3.55 + 11.46
```
     
Model for school 7930:

MathScores = 38.75 + 7.91*timeonmath

>Q: What is the unique intercept for school 7472?

A: 50.68 (it’s the reference)
    

>Q:  What is the unique slope for school 7472?

A: -3.55 (again, reference)   

     
     

Now we could do this all day and enter them one at a time, but that would be difficult and probably lead to an error or two... So we should probably do it in R, which is much easier and less likely to lead to an error.

##### Getting intercepts and Slopes in R

Remember that what we want to end up with is a dataframe with a row per school and at least 3 columns:

1. schoolid
2. intercept estimates
3. slope estimates

In addition to these three variables, we'd want some predictor(s). In this case, we'll use `schooltype`, which is a code for whether schools are public or private (in the nels88 data). We'll start by getting the df with just the first three columns and then add in the predictor variable.

The code we'll use is admittedly a little complicated at first glance, but basically all we're doing is:

1. Splitting the df into a df per school
2. Regressing mathscore on timeonmath in each df
3. Returning a df of coefficients (one intercept and one slope per school)
4. Labeling the intercept and slope
5. Tidying the data by:
    1. Turning schoolid back into a column (currently spread across the data as column names)
    2. Spreading the coefficients so that we have separate columns for the intercept and slope.

```{r get_ints_and_slopes}
ints_slopes <- nels_88 %>% 
  split(.$schoolid) %>% # split df into list of dfs, one per school
  map(~ lm(mathscore ~ timeonmath, data = .)) %>% # run OLS predicting score from time
  map_dfc("coefficients") %>% # get coefficients for each in a df
  cbind(term = c("intercept", "b_timeonmath"), .) %>% # identify the row with intercept and slope
  gather(schoolid, estimate, -term) %>% # gather
  spread(term, estimate)     # spread so we have a row per school, 
                             # and a column for int and slope

ints_slopes
```

### Part 2: Slopes and Intercepts as Outcomes

#### Step 1: Create Dataset with Slopes, Intercepts, and Group-level Predictors

Now we just need to add the `schooltype` column from the `nels_88` dataframe. We can use one of the `join` functions from the `dplyr` library. This will use the schoolid variable as the *key* to match up the data from the two dataframes.

First, though, we'll select just the schoolid and schooltype variables from nels_88 since that is all we need now. The last thing we'll do is run `distinct()` on it to remove the repeats; without this, there would be a row per student and we just want a row per school.

```{r}
ints_slopes <- nels_88 %>% # start with nels data
  select(schoolid, schooltype) %>% # select just schoolid (for matching) & schooltype
  left_join(ints_slopes) %>%  # left join
  distinct() # get just distinct (not repeated) rows

ints_slopes
```

#### Regress Intercepts on Predictor(s)

Now we can see if private and public schools have different intercepts by regressing the intercepts on the schooltype predictor variable.

```{r schooltype_math0}
ints_model <- lm(intercept ~ schooltype, data = ints_slopes)
summary(ints_model)
```

>Q:  What is the regression equation predicting intercept from school type?    

Schooltype: 0=private, 1=public

A: Intercept = 59.21 - 16.06*schooltype       

>Q: What does the intercept mean?    

A: b0 = 59.21, Predicted intercept when schooltype = 0, meaning the predicted  intercept for private schools     

>Q:  What does the slope for schooltype mean (using the number in your answer)?    

A: b1 = -16.06, the predicted change in the intercept when schooltype = 1, meaning the predicted difference in the intercept for public schools compared to private schools


>Q: What are the expected intercepts for private and public schools?    

A: Private schools: predicted intercept = 59.21
  Public schools: predicted intercept = 59.21-16.06 = 43.15
```{r}
59.21-16.06
```


#### Regress Slope on Predictor(s)

Now we can see if private and public schools have different slopes for time spent on math by regressing that slope on the schooltype predictor variable.

```{r schooltype_mathtime}
slopes_model <- lm(b_timeonmath ~ schooltype, data = ints_slopes)
summary(slopes_model)
```

>Q:  What is the regression equation predicting the slope from school type?    

0=private, 1=public

A: Slopes = 1.09 + 0.96*schooltype

>Q: What does the intercept row mean?    

A: b0 = 1.09, predicted slope for private schools          

>Q:  What does the slope for schooltype mean (using the number in your answer)?    

A: b1 = 0.96, the amount that we expect the predicted slope to change by for public schools compared to private schools        

>Q: What are the expected slopes for private and public schools?    

A: Predicted slope for private schools = 1.09
  Predicted slope for public schools = 1.09 + 0.96 = 2.05
```{r}
1.09+0.96
```
  

>Q: Is the relationship between time spent on math HW and math score significantly different in private vs. public schools?    

A: No

We could visualize this in ggplot by mapping the color aesthetic onto schooltype, like so:

```{r schootype_plot}
ggplot(data = nels_88, aes(x = timeonmath, y = mathscore, 
                           color = schooltype, group = schoolid)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) + 
  theme_minimal()
```

So it looks like we had just 1 private school, so this analysis is not a very good test of public vs. private school differences.


# Random Coefficients Regression: "Full" MLM

## Conceptual Review

Now we'll run a "full" random coefficients regression. In RCR, every L1 coefficient (the betas) is allowed to vary randomly at the group level (L2). 

>Q: What would the RCR model be to predict math scores from hours spent on math homework, school type (public vs. private), and the interaction between time on math and school type?

### Notation

#### Multiple Equation Form:

L1: 	

$$Y_{ij} = b_{0j} + b_{1j}time_{ij} + e_{ij}$$

L2: 	

$$b_{0j} = g_{00} + g_{01}type_j + u_{0j}$$
$$b_{1j} = g_{10} + g_{11}type_j + u_{1j}$$

#### Single Equation Form

$$Y_{ij} = g_{00} + g_{01}type_j + g_{10}time_{ij} + g_{11}time_{ij}*type_j + (u_{0j} + u_{1j}time_{ij} + e_{ij})$$

What is the meaning of each parameter?

$g_{00}$ = *expected MathScore when time spent on HW = 0 for students in private schools (when type = 0); overall intercept*
				
$g_{01}$ = *main effect of L2 predictor (type); difference in intercept for private vs. public; difference between expected MathScore when time spent on HW=0 for people in public schools versus private schools *
				
$u_{0j}$ = *random error for the intercept. In random coefficient regression, we estimate var(u0)=tau00 Since we have a L2 predictor, this tells us how much residual (school-to-school) variance there is in the intercepts after accounting for the effect of our L2 predictor. If we're doing a good job explaining variance this will be low (might have reduced for example after adding school).*

$g_{10}$ = *In other words, the main effect of time spent on HW when Type=0. If your L2 variable were continuous, this would still be the main effect of time spent on HW when your L2 variable=0. In this case, it is expected increase in MathScore for each 1 unit increase in time spent on HW for private schools.*

$g_{11}$ = *effect of Type on the relationship between MathScore and time spent on HW (cross-level interaction). In other words, the effect of being in a private or public school on the slope predicting MathScore from Time spent on HW (effect of school type on the relationship between time and score)*

$u_{1j}$ = *random error for the slope. In random coefficient regression, we estimate* $var(u_{1j}) = \tau_{11}$. *Since we have a L2 predictor for the slope, var(u1j) is how much residual variance there is in the slopes after taking into account our L2 predictor. *

$e_{ij}$ = *residual variance at the individual level (error)*

>Q: What if you found that the slopes did NOT vary randomly across groups? How would that model be different?

A: *In this model, since we have a L2 predictor for our slope equation, that would mean the random variance in slopes is completely explained by our L2 predictor. This means that the slope for all private schools is the same and the slope for all public schools is the same. When u1j is in our model, that means that the effect of being public or private on slopes can vary from school to school. So, this would mean that u1j = 0.*

## MLM in R

We'll be using `lmer` from the `lme4` library to run the Random Coefficients Regression. Y'all might recall running some `lmer` models for random effects anova in 612, and you'll hopefully recognize the syntax.

### NULL Model & ICC

First, it's common to estimate a NULL model (i.e., a model with just a fixed and random intercept). This can be used as a baseline in model comparisons and, maybe more importantly, can be used to get the ICC.

>Q: What does the intercept only equation look like?

A: 

$$L1: Y_{ij} = b_{0j} + e_{ij}$$
$$L2: 	b_{0j} = g_{00} + u_{0j}$$

Remember, in `lme4` we use the `lm()` syntax, but add the `| GROUP` syntax to specify random effects. So, for the intercept only, we do this:

Multilevel modeling notation:

* lmer(DV ~ IV1 + IV2 + ... etc, (random intercept and/or slope terms | Grouping Variable(s)))
```{r lmer_null}
rcr_0 <- lmer(mathscore ~ 1 + (1 | schoolid),
              data = nels_88)
summary(rcr_0)
```

>Q: How can we calculate the ICC from this output?    

A: ICC = between-group variance / total variance
```{r}
34.01/(34.01 + 72.26)
```

  

>Q: What does this mean?    

A: 32% of the variance in math scores is accounted for by differences among schools.
    
    

#### Calculating ICC in R

We can calculate this in R by running VarCorr on the model, which provides the random effects parameters as variance-covariances (and SD-Correlations). We'll do that and a little data manipulation to calculate ICC.

```{r get_icc}
variances <- as.data.frame(VarCorr(rcr_0))
variances

ICC <- variances[1,4]/(variances[1,4] + variances[2,4])
ICC
```

Now we'll look at the full model.

### Full Model

Let's predict mathscores from time spent on math homework, school type (private vs. public), and the cross-level interaction between these variables.

```{r}
rcr_1 <-  lmer(mathscore ~ 1 + timeonmath*schooltype + (1 + timeonmath | schoolid),
              data = nels_88)


summary(rcr_1)
```

>Q: What is the regression equation?   

A:

$$Y_{ij} = g_{00} + g_{01}type_j + g_{10}time_{ij} + g_{11}time_{ij}*type_j + (u_{0j} + u_{1j}time_{ij} + e_{ij})$$

$$Y_{ij} = 59.21 - 15.97 (type) + 1.09 (time) + 0.95 (time*type) + error$$
*Note that the error terms are going to differ for each person in each school. Remember that the output gives you error variance not actual error values.*

>Q: What do the fixed effects (including the intercept) mean?     

A:

Intercept: 59.21, predicted math score when timeonmath = 0 and type = 0 (in private schools)

g01 (type): -15.97, predicted change in the model intercept when type = 1 (in public schools)

g10 (time): 1.09, slope representing the relationship between timeonmath and predicted math scores when type = 0 (in private schools)

g11 (time*type): 0.95, predicted change in the model slope for when type = 1 (in public schools)
    

>Q: What about the random effects?     

A: var(u0j) = 51.84
   var(u1j) = 27.27
   var(eij) = 42.96
    

### Comparing Models

Remember: you can compare most kinds of models in R with the `anova()` function. We can compare these models in terms of their deviance, the difference of which is $\chi^2$ distributed (for a significance test).

```{r}
anova(rcr_0, rcr_1)
```

>Q: What does this comparison tell us?    

A:  The full model is a significant improvement to the null model. 

And you can compare models with and without random or fixed effects. For example, let's see how our full model compares to one without a random slope for time on math.

```{r}
rcr_2 <- lmer(mathscore ~ 1 + timeonmath*schooltype + (1 | schoolid),
              data = nels_88)
anova(rcr_1,
      rcr_2)

```

>Q: What does this comparison tell us?    

A: The full model containing the random slopes term accounts for significant more variability in math scores than the model that does not contain the random slopes term. 


## Sample Write-up

In this study, we investigated the relationship between time spent on math homework and math scores among students from ten different schools. Schools accounted for a substantial amount of variability in math scores, ICC = .32. Therefore, we conducted a multilevel model with students nested in schools.

I ran a random coefficients regression in which students were nested in schools and both intercepts and slopes were allowed to vary randomly. I predicted students' math scores from the time they spent on their math homework, the type of school (public vs. private) that they attended, and the interaction between time spent on homework and type of school. 

Time spent on math homework did not significantly predict math scores for private schools (*t*(7.40) = 0.21, *p* = .841). Additionally, the relationship between time spent on math homework and math scores did not significantly differ between private and public schools, (*t*(7.17) = 0.95, *p* = .869).


