---
title: 'Lecture 5: Interactions'
author: "ETB"
date: "4/12/2022"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(foreign)
```

## Importing data from SPSS

Sadly, a lot of data files will be in the lame SPSS ".sav" format. Happily, we can import those using the package "foreign". The key function here is read.spss() and note the "to.data.frame" argument:
```{r import from SPSS}

setwd("/Users/Berkman/Documents/Psychology/Oregon/Teaching/Multivariate/Spr 2022/Lecture data")
dataset = read.spss("Lecture5NEO.sav", to.data.frame=TRUE)

# Now grab just the three variables we'll need for this lecture
keyVars = dataset[,c("soitot", "sos", "a")]
```

## Create the variables we'll want

Let's create the uncentered interaction term and the centered variables, then look at the correlation matrix
```{r create vars}
attach(keyVars)
keyVars[,"sosXa"] = sos*a
keyVars[,"c_sos"] = sos-mean(sos)
keyVars[,"c_a"] = a - mean(a)
attach(keyVars)

#Look at the correlation matrix
cor(keyVars)
```

## Uncentered model with only main effects (no interaction)

Here's a model with only main effects. Since we're not looking at the interaction, we don't need to center anything
```{r uncentered MEs only}
me_model = lm(soitot ~ c_sos + c_a) 
summary(me_model)
```

## Uncentered model with main effects + an interaction

Now what happens if we run an interaction without centering?
```{r uncentered full model}
int_model = lm(soitot ~ sos + a + sos*a)
summary(int_model)
```

## Centered model with main effects + an interaction

And what happens if we run that same model but center first?
```{r centered full}

# Build data frame for the centered model
KeyVarsCent = dataset[,c("c_sos", "c_a")]
KeyVarsCent[,"sosXa_cent"] = c_sos*c_a 
attach(KeyVarsCent)

# Centered model with MEs and an interaction
full_model = lm(soitot ~ c_sos + c_a + sosXa_cent)
full_model = lm(soitot ~ c_sos*c_a)
summary(full_model)
summary(full_model)$cov.unscaled # Get coefficient correlations
```

## Plotting the simple slopes

First, create a variable to represent the range of sexual attitudes along the x-axis
```{r x axis}
SexAtt <- seq(20, 120, by=.1) - mean(sos)  # Mean sexual attitudes = 73.62
```

Now we need to get the "simple" intercepts and slopes from the model we used above. The overall equation is:
SOI = b0 + b1\*sos + b2\*a + b3\*sos\*a

So, if we're treating *SOS* continuous and plotting it along the x-axis, then we need to look at high, medium, and low agreeableness. So, we'll plug in discrete values for "a" where it appears in that equation, turning the "a" into constants. We can re-write the equation as:

SOI_at_a = (b0 + b2\*a) + (b1\*sos + b3\*sos\*a)

To make use of that, let's get the parameter values:
```{r pull parameters for 2-way}
b0 = full_model$coefficients[1]
b1 = full_model$coefficients[2]
b2 = full_model$coefficients[3]
b3 = full_model$coefficients[4]
```

Next, create three "simple lines" to plot on the y-axis using the slopes and intercepts calculated in our equation:
```{r simple lines for 2-way}

BehHighA <- (b0 + b2*sd(a)) + (b1*SexAtt + b3*SexAtt*sd(a))  # -.158 + SexAtt*.018
BehMeanA <- (b0 + b2*0) + (b1*SexAtt + b3*SexAtt*0)  # -.032+SexAtt*.024
BehLowA <-  (b0 - b2*sd(a)) + (b1*SexAtt - b3*SexAtt*sd(a))   # .094+SexAtt*.03
```

Now we can plot the first simple line:
```{r plot first line}
plot(SexAtt,BehLowA,type="l", col="green", lwd=3, xlab="Sexual attitudes (centered; higher numbers = more liberal)", ylab="Sexual behavior (centered; higher numbers = more active)")

# And also plot the other two lines using the lines() function. (We use this instead of plot() so not to overwrite the original graph)

# Plot the other two lines (using "lines" instead of "plot" so not to overwrite)
lines(SexAtt,BehMeanA,type="l", col="black", lwd=6)
lines(SexAtt,BehHighA,type="l", col="red", lwd=3)

#Add a legend at x=15, y=-0.5

legend(15,-.5, c("Low A", "Mean A", "High A"),
       lty=c(1,1,1),
       lwd=c(3,6,3),
       col=c("green","black","red")
)
```

## The faculty salary example that we walked through in class

Import data from SPSS, use "to.data.frame=TRUE":
```{r pull in data}
# Salary dataset from class
dataset = read.spss("Lecture6Salary.sav", to.data.frame=TRUE)
attach(dataset)
```

## Main effects of publication and sex, and their interaction

Run the full model:
```{r full model w/ interactions}

full_model = lm(salary ~ c_pubs + female + c_pubs*female, data = dataset)
summary(full_model)

# recode the dummy variable
dataset$female <- relevel(female, ref="Female", data=dataset)

```



## Separate by gender
Now let's look at this same analysis separately for two sexes (i.e., simple slopes/intercepts)
```{r separte models}
mdata = subset(dataset,female=="Male")
fdata = subset(dataset,female=="Female")

# Run models for each sex separately to test simple effects
m_model = lm(salary ~ c_pubs, data = mdata) # Males
anova(m_model)

f_model = lm(salary ~ c_pubs, data = fdata) # Females
anova(f_model)
```

## Simple slope: Recode method
Now let's test the simple slope of pubs for women by recoding the "female" variable with women as the reference
```{r recode method}

relevel(dataset$female, ref = 'Female')

full_model_relevel = lm(salary ~ c_pubs + female + c_pubs*female, data = dataset)
summary(full_model_relevel)

```


## Three-way interactions
What if we wanted to lok at the three-way interaction, where salary is a function of pubs and sex and also cites?

Well, first run the model:
```{r threeway}
three_way = lm(salary ~ c_pubs*c_cits*female)

summary(three_way)
```
What you get from this is *8* parameters: an intercept, three main effects, three two-way interactions, and one three-way interaction.

It looks like there are effects of pubs and cites but no two- or three-way interactions. Still, let's visualize for fun.

To do this, we need to get the "simple" intercepts and slopes from the model we used above. The overall equation is:
Salary = b0 + b1\*pubs + b2\*cites + b3\*female + b4\*pubs\*cites + b5\*pubs\*female + b6\*cites\*female + b7\*pubs\*cites\*female

So, if we're treating *pubs* as continuous and plotting it along the x-axis, then we need to look at high, medium, and low *cites*, separately for men and women. So, we'll plug in discrete values for "cites" and "female" where they appear in that equation, turning the "cites" and "female" values into constants. We can re-write the equation as:

Salary_at_cites_and_female = (b0 + b2\*cites + b3\*female + b6\*cites\*female) + pubs\*(b1 + b4\*cites + b5\*female + b7\*cites\*female)

To make use of that, let's get the parameter values:
```{r pull parameters for three-way}
b0 = three_way$coefficients[1]
b1 = three_way$coefficients[2]
b2 = three_way$coefficients[3]
b3 = three_way$coefficients[4]
b4 = three_way$coefficients[5]
b5 = three_way$coefficients[6]
b6 = three_way$coefficients[7]
b7 = three_way$coefficients[8]
```

Next, create six "simple lines" to plot on the y-axis using the slopes and intercepts calculated in our equation:
```{r simple lines for 3-way}

pubs_range <- seq(min(c_pubs), max(c_pubs), by=.1)

# Lines for MEN (i.e., female = 0)
HighCitesM <- (b0 + b2*sd(c_cits)) + pubs_range*(b1 + b4*sd(c_cits))
MeanCitesM <- (b0 + b2*0) + pubs_range*(b1 + b4*0)
LowCitesM <-  (b0 - b2*sd(c_cits)) + pubs_range*(b1 - b4*sd(c_cits))

# Lines for WOMEN (i.e., female = 1)
HighCitesW <- (b0 + b2*sd(c_cits) + b3*1 + b6*sd(c_cits)*1) + pubs_range*(b1 + b4*sd(c_cits) + b5*1 + b7*sd(c_cits)*1)
MeanCitesW <-(b0 + b2*0 + b3*1 + b6*0*1) + pubs_range*(b1 + b4*0 + b5*1 + b7*0*1)
LowCitesW <- (b0 - b2*sd(c_cits) + b3*1 - b6*sd(c_cits)*1) + pubs_range*(b1 - b4*sd(c_cits) + b5*1 - b7*sd(c_cits)*1)

```

Plot the results separately for M and W using the par() function
```{r plot}
par(mfrow=c(1,2))

plot(pubs_range,LowCitesM,type="l", col="green", lwd=3, xlab="Publications (centered)", ylab="Salary", ylim=c(40000,100000))
title("Salary as a function of publications and salary for MALES")
# And also plot the other two lines using the lines() function. (We use this instead of plot() so not to overwrite the original graph)
# Plot the other two lines (using "lines" instead of "plot" so not to overwrite)
lines(pubs_range,MeanCitesM,type="l", col="black", lwd=6)
lines(pubs_range,HighCitesM,type="l", col="red", lwd=3)

#Add a legend at x=15, y=-0.5
legend(-10,100000, c("Low Cites", "Mean Cites", "High Cites"),
       lty=c(1,1,1),
       lwd=c(3,6,3),
       col=c("green","black","red")
)

plot(pubs_range,LowCitesW,type="l", col="green", lwd=3, xlab="Publications (centered)", ylab="Salary", ylim=c(40000,100000))
title("Salary as a function of publications and salary for FEMALES")
# And also plot the other two lines using the lines() function. (We use this instead of plot() so not to overwrite the original graph)
# Plot the other two lines (using "lines" instead of "plot" so not to overwrite)
lines(pubs_range,MeanCitesW,type="l", col="black", lwd=6)
lines(pubs_range,HighCitesW,type="l", col="red", lwd=3)

#Add a legend at x=15, y=-0.5
legend(-10,100000, c("Low Cites", "Mean Cites", "High Cites"),
       lty=c(1,1,1),
       lwd=c(3,6,3),
       col=c("green","black","red")
)

```