---
title: "Lecture 2a"
author: "Janette Avelar"
date: "4/5/2022"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(pracma)
library(car)
library(MASS)
```

# Week 2 - 4/5/22

*Programming in R Part I: Intro and Matrix Manipulation*

1. How do you create a matrix?

```{r creating a matrix}
x = matrix(c(2, 1, -1, -1, -1, 1, -1, -5, -5, 10, 2, 2, 2, 2, -8),
           byrow = TRUE, ncol = 5)
print(x)
```

Now that you have a matrix, you can use indexing to call up specific elements in the matrix. For instance, what's the element in the 3rd row and 4th column?

```{r}
x[3,4]
```

2. Getting help
?function = search help; sometimes doesn't work
??function = pulls up all documentation
?? "function" = searches for the term within ""; useful when you don't know what function you need to do something

3. Reduced row echelon form

```{r reduced row echelon form}
rref(x)
# you can also do reduced row form for the transpose
rref(t(x))
```

## Random Numbers AKA Simulation

Something really useful is to generate datasets to work with.
You need random numbers for this.

*And* you need to keep in mind what type of distribution you want.

A uniform distribution

```{r uniform distribution}
runif(1) # stands for random uniform
# () specifies numbers you want
# defaults to a random number from 0 to 1

runif(2, min = 3, max = 5) # generates 2 random numbers from a uniform distribution 3:5

hist(runif(10000000)) # wow it's a rectangle ~uniform~
```

What about a normal distribution?

```{r normal distribution}
rnorm(1) # stands for normal distribution
# () specifies numbers you want
# defaults mean=0, SD=1
```

Note that in these functions each random sample is independent of one another--so you can get duplicates.

What if we don't want to draw randomly this way and we want whole numbers? For example, if we're pulling from a deck of cards.

```{r sampling}
sample(1:10, 3) # defaults to no replacement (meaning you can run out)
sample(1:10, 3, replace = TRUE) # with replacement
# note if you keep running that you might get a couple duplicates

sample(1:5, 300, replace = TRUE) # more likely to get duplicates because it's a bigger sample
# note this could simulate survey responses where 1:5 stands in for likert scale responses
```

Now you can use these functions to create a random matrix

```{r}
x <- matrix(runif(100), ncol = 1) # [100,1] from uniform distribution

x <- rnorm(5000) # 5000 numbers from normal distribution
mean(x) # should be close to 0

x <- rnorm(1, mean = 100, sd = 15) # draw 1 number from normal distribution w/ mean = 100 and SD = 15
```

Reminder for simulations that the bigger your sample gets, the closer to 0 the variance.

## Flow Control

You can do amazing things with these key pieces:
* if/else statements
* for loops
* logical (boolean) operators

### Boolean operators

```{r logic}
a = 3 # this is ok
3 = a # this is not

a <- 3 # 2 key strokes vs. 1... why not just use =
4 -> a # r will accept this; but don't do it

a == 3
a == 4
# vs.
a != 4
a != 3
```

Where do boolean operators become useful?

```{r boolean example}
y = 1:10 # create matrix with numbers 1 through 10
y > 4 # tells you which elements are greater than 4
y[y > 4] # pulls up elements greater than 4 rather than returning logic
# useful if you want to create a new variable e.g.
```

### If/Else statements

Basic structure:
  if(LOGICAL CONDITION IS TRUE) {
    DO SOMETHING  }
  elseif (OTHER LOGICAL CONDITION) {
    SOMETHING ELSE    }
  print(WHAT YOU WANT IT TO SAY)
    
```{r if/else example}
x <- runif(1)

if(x < (1/3)) {
  print("It's a one!")
} else if (x < (2/3)) {
  print("It's a two!")
} else {
  print("It's a three!")
}
```

For loops: do the same operation to a variable a fixed number of times

Basic structure:

    for (VARIABLE IN VECTOR) {
      do this thing where VARIABLE changes each iteration through the loop
    }
    
```{r for loop example}
# simple
for (i in 1:5) {print(paste("i", i))}
# 1-5 is a list of numbers, so what this for loop does is takes the value of
# each element and prints them

# another simple
for(i in 1:100){print(rnorm(1))}
# for a set of numbers between 1 and 100 in a normal distribution
# pull out 100 random numbers

# remember the (i in x) part, is you specifying the elements in a given matrix

x = 0
for(i in 1:10) {
  x = x + i
  print(x)
}
```

One last example, say you ran a trial with 7 trials You're now wanting to clean up your data.

First, you'll create your data variable, and another object specifying the dimensions of your data (7 since you had 7 trials).

We then want to build a for loop that will determine if the data is NA or not. If it's NA we want to take the value and replace it with -1.

```{r applied for loop example}
data = c(6, NA, 3, 7, 3, NA, 8)
n_trials = length(data)

for(trial in 1:n_trials) {
  if(is.na(data[trial])) {
      data[trial] = -1
  }
}
data # NAs are now -1
```

Here's a simpler way to do this.

```{r simpler applied example}
data = c(6, NA, 3, 7, 3, NA, 8)
```

## Nested for loops

You can put loops inside of your loops!

First, create a matrix

```{r}
x <- zeros(3, 4) # creates 3x4 empty matrix of zeros
```

Now we can specify our loops. Let's say we want to create a matrix where r (rows) 1-3 and c (cols) 1-4

```{r}
for(r in 1:3) { # for every row 1-3, starting with 1st row
  for(c in 1:4) { # go col by col (in that row) starting with 1
    x[r,c] = r * 10 + c # the value that goes into that spot is rows + columns
  }
}

x # now you've filled in x which was previously empty
```

## Practical Applications of For Loops

Start by simulating some real statistical data.

```{r}
for(n in c(5, 10, 20, 100, 10000)) { #use these values (not the element place like before)
  x = rnorm(n) 
  print(sd(x)) 
}
```

## Example Simulation 1

I have heard that we can use the binomial distribution to figure out probabilities of count variables. For example, if I want to know the likelihood of getting exactly 8 heads out of 10 coin flips, I can look it up in the binomial probability distrubution function, which simply follows the formula.

```{r}
dbinom(8, 10, 0.5) # If I flip a thing with a .5 probability 10 times, how many times will it be 8?
```

But suppose I don't trust the dbinom() function for wahtever reason, and I want to check it myself. How could I do that with a simulation that flips a coin and counts for us?

Step 1: Flip a coin 10 times. We can use sample() for this, where 1 = heads and 0 = tails

```{r}
one_try <- sample(0:1, 10, replace = TRUE)
one_try
```

Step 2: Then I want to count how many are heads. You could count the "1s", or just sum up the sample

```{r}
nHeads <- sum(one_try)
nHeads
```

Step 3: Write that number down, do it some more and keep track of it. We can use a for loop!

```{r}
for(samples in 1:1000000) {
  one_try = sample(0:1, 10, replace = TRUE)
  nHeads = sum(one_try)
}
#where's our info? Don't forget you need to ask it to print by ssaving the output!

# We need a storage vector.
```

Step 4: Do that again, but with an empty storage vector.

```{r}
storage <- vector() # or c()

for(samples in 1:1000000) {
  
  # first flip 10 coins
  one_try = sample(0:1, 10, replace = TRUE)
  
  # second count how many heads there are
  nHeads = sum(one_try)
  
  # third put sum of one_try into our storage vector
  storage[samples] = nHeads
}

# but we still haven't asked it to print
hist(storage)
```

What percentage of the time was it exactly 8 out of the million flips?

```{r}
# dumb way to do it but just for practicing loops
nEights = 0

for(headCounts in 1:1000000){
  if(storage[headCounts]==8) {nEights = nEights+1}
  else{nEights = nEights}
}

NEights

# more simple way
nEights = sum(storage == 8)
nEights
```

