---
title: 'Lecture 11: Path Analysis'
author: "ETB"
date: "5/2/2022"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(lavaan)
require(semPlot)
require(foreign)
require(tidySEM)

setwd("/Users/Berkman/Documents/Psychology/Oregon/Teaching/Multivariate/Spr 2022/Lecture data")
data <- read.csv("Lecture11_Path.csv")

```

## Take a look at the data
```{r Examine some data}

# Select out just the variables we want
selectVars <- c("momalc3", "momcig3", "gender", "eth", "hcirc0", "hcirc36")
data_path <- data[selectVars]

# Clean up, recode as factor
data_path[data_path==999] = NA
data_path$gender <- factor(data_path$gender, labels = c("female", "male"))
data_path$eth <- factor(data_path$eth, labels = c("white", "Black"))

# Take a look
summary(data_path)
cor(data_path[c("momalc3", "momcig3", "hcirc0", "hcirc36")], use = 'pairwise.complete.obs')
```

## Build and fit the model

Here's how you build a basic path model in lavaan as shown in the lecture notes. Note that Lavaan likes characters as inputs (in single quotes)
```{r define model}
model_pathExample <- '

hcirc36 ~ hcirc0 + gender + eth
hcirc0 ~ momalc3 + momcig3 + gender + eth

'
```

Then run the model with the cfa() function and look at the solution with summary()
```{r run and inspect}

fit_pathExample <- sem(model_pathExample, data = data_path)
summary(fit_pathExample, fit.measures = TRUE)

```

## Standardized model and plot

Use the handy functions to look at the standardized solutions (i.e., in z units) and the residuals
```{r inspect more}

standardizedsolution(fit_pathExample)  # standardized solution
# resid(fit_pathExample)  # Check the residuals

```

And, of course, you can always just plot the data
```{r plot}

layout_path <- get_layout("momalc3", "hcirc0", "hcirc36",
                     "momcig3", "", "", 
                     "gender", "", "",
                     "eth","","", rows = 4)

graph_sem(model = fit_pathExample,layout = layout_path)


```