---
title: 'Lecture 13: Logistic Analysis'
author: "ETB"
date: "5/10/2022"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    df_print: paged
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#rm(list=ls())
library(ggplot2)
library(knitr)
library(here)
library(rio)
library(tidyverse)
# setwd('~/Documents/Psychology/Oregon/Teaching/Multivariate/Spr 2022/Lecture Data/')
```

## Bring in the data

```{r import data}
# Get the data (note that it's in a CSV today)
data.logit <- import(here("Data", "Lecture13IPV.csv"),
                     setclass = "tibble")
#look at our data to get familiar
head(data.logit)
#note what we're looking at is reasons people give for being in a relationship
#in order to predict likelihood of intimate partner violence

# Convert the outcome to a factor
data.logit$AnyPhys = factor(data.logit$AnyPhys) #any physical aggression
```

## Run the actual model

```{r run the logit model and inspect it}

# Run the model using the "glm()" function
logitModel = glm(AnyPhys ~ EXTSCORE, #EXTSCORE is extrinsic motivation
                 data = data.logit, 
                 family= binomial(link="logit"), 
                 na.action=na.omit) 

# Summary. Remember that the bs are LOG ODDS
summary(logitModel)
#Null Deviance refers to model with no predictors
#Residual Deviance includes predictor
#note the change in deviance score; it's indicative of what we WANT
#we want it to go down when we add a predictor
#note you can use a deviance change test (chi square distributed)
anova(logitModel, test = "Chisq")#passing through ANOVA does the changes test for you
#=model comparison

# Exponent of the model to get the odds
exp(coef(logitModel))

```

## Classification table

```{r fitted values}

# Fitted values and thresold
Yhat <- fitted(logitModel)
thresh  <- 0.5

# Dichotomize accordint to predicted values and threshold
YhatFac <- cut(Yhat, breaks=c(-Inf, thresh, Inf), labels=c("Predicted No", "Predicted Yes"))

# Make a contingency table, omitting cases where EXTSCORE has missing data
levels(data.logit$AnyPhys) <- c("Absent", "Present")
cTab <- table(subset(data.logit$AnyPhys, data.logit$EXTSCORE!="NA"), YhatFac)
# Does anyone know why this table is so ugly?
kable(cTab, format="markdown", align="c", caption="Predicted versus actual outcomes")

# percentage correct for training data
sum(diag(cTab)) / length(subset(data.logit$AnyPhys, data.logit$EXTSCORE!="NA"))
```

## Plotting

The following is from Stack Overflow: http://stackoverflow.com/questions/26694931/how-to-plot-logit-and-probit-in-ggplot2. It's a nice example of plotting the (fitted) model along with the raw data points. It involves creating predicted values based on the model output (using the "predict" function) then plotting the results of that model. 

```{r plot single predictor LR}

# Create a temporary data frame of hypothetical values
# Create a variable to represent "EXTSCORE" along the x-axis
temp.data <- data.frame(EXTSCORE = seq(-55,18,.01))

# Predict the fitted values given the model and hypothetical data
predicted.data <- as.data.frame(predict(logitModel, newdata = temp.data, type="link", se=TRUE))

# Combine the hypothetical data and predicted values
new.data <- cbind(temp.data, predicted.data)

# Calculate confidence intervals and put them in the "new.data" frame
std <- qnorm(0.95 / 2 + 0.5)
new.data$ymin <- logitModel$family$linkinv(new.data$fit - std * new.data$se)
new.data$ymax <- logitModel$family$linkinv(new.data$fit + std * new.data$se)
new.data$fit <- logitModel$family$linkinv(new.data$fit)  # Rescale to 0-1

# Plot the actual data (in "data.logit") and the fitted data (in "new.data")
p <- ggplot(data.logit, aes(x=EXTSCORE, y=as.numeric(AnyPhys)-1))
p + geom_point() +
    geom_ribbon(data=new.data, aes(y=fit, ymin=ymin, ymax=ymax), alpha=0.5) +
    geom_line(data=new.data, aes(y=fit)) + 
    labs(x="Extrinsic Relationsip Motivation", y = "P(Physical Aggression)")

```


## Hierarchical model

```{r Hierarchical (stepwise) model}
logitModel2 = glm(AnyPhys ~ CHILDS + EXTSCORE, data = data.logit, family = "binomial")
logitModel3 = glm(AnyPhys ~ CHILDS * EXTSCORE, data = data.logit, family = "binomial")
summary(logitModel2)
exp(coef(logitModel2))

# Compare to previous model
anova(logitModel, logitModel2, logitModel3, test = "Chisq")

## Add the interaction
logitModel3 = glm(AnyPhys ~ c_Ext*c_childs, data = data.logit, family = "binomial")
summary(logitModel3)

mean(data.logit$c_childs)

# Compare the previous models
anova(logitModel2, logitModel3, test = "Chisq")
```


## Plotting logistic regression

```{r plot the relationship in probability terms}
# Create a variable to represent "EXTSCORE" along the x-axis
ext <- seq(-55,18,.01)

# P(violence) as a function of EXTSCORE
# b0 = -.086
# b1 = .032
p_v <- exp(-.086 + ext*.032) / (1+exp(-.086 + ext*.032))

# Plot ext (on x) by p_v (on y) in black
plot(ext,p_v, type="l", col="black", xlab="Extrinsic motivation score", ylab="P(violence)", main="Probability of committing violence as a function of EXT score")

# Plot a dashed red horizontal line along the range of extscore at p = .05
size_ext <- length(ext)
lines(ext,rep(0.5, size_ext), type="l", col="red", lty=2)

# Plot a dashed blue vertical line along the range of P(violence) at the threshold
# Threshold = -b0 / b1
thresh <- .086/.032
size_p_v = length(p_v)
lines(rep(thresh,size_p_v),seq(0,1,len=size_p_v), type="l", col="blue", lty=2)

# Plot the interaction between "ext" and "childs" in predicting violence
ext_c <- seq(-35, 38, .01)
high_childs <- 0.12 + 0.055*ext_c
avg_childs <- -0.852 + 0.023*ext_c
low_childs <- -1.82 - 0.009*ext_c

p_pa_high_childs <- exp(high_childs) / (1+exp(high_childs))
p_pa_avg_childs <- exp(avg_childs) / (1+exp(avg_childs))
p_pa_low_childs <- exp(low_childs) / (1+exp(low_childs))

plot(ext_c, p_pa_high_childs, type="l", col="blue", xlab = "Extrinsic motivation score (centered)", ylab="P(violence | ext, childs)", main="Probability of committing violence as a function of EXT score")
lines(ext_c, p_pa_avg_childs, type="l", col="green")
lines(ext_c, p_pa_low_childs, type="l", col="red")
legend(-35, 0.9, c("Probability of PA for High childish behavior", "Probability of PA for average childish behavior", "Probability of PA for low childish behavior"), col=c("blue", "green", "red"), lty=c(1,1,1))
```

