---
title: "Problem Set 3"
author: "Janette Avelar"
date: "4/27/2022"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 999)
```

```{r library, include=FALSE}
library(rio)
library(here)
library(psych)
library(OpenMx)
library(tidyverse)
library(lavaan)
# library(semPlot)

q1data <- import(here("data", "PS3_Q1-3.RData"),
                 setclass = "tibble")
q4data <- import(here("data", "PS3_Q4.RData"),
                 setclass = "tibble")
```

## Question 1

**Run a components analysis on all 19 items. Be sure to generate the reproduced correlation matrix. Use principal components extraction to extract any factors with an eigenvalue greater than 1, and** *varimax* **orthogonal rotation.**  

This is the original correlation matrix with heatmap visualization:

```{r og cor mat}
# get a correlation matrix
# focus on 19 items--therefore, we'll drop first 2 cols
q1data_mod <- q1data %>% 
  select(c(3:21))
# create cor
og_cor <- data.frame(cor(q1data_mod, use = "pairwise.complete.obs"))

round(og_cor, 2)

# create a heatmap to visualize

## convert the correlation matrix to long format
og_cor_long <- og_cor %>%
  rownames_to_column(var = "var1") %>%
  gather(key = "var2", value = "corr", -var1)

# plot 
ggplot(data = og_cor_long, aes(x = var1, y = var2, fill = corr)) + 
  geom_tile()
```

Next, we'll run an unrotated PCA and extract eigenvalues.

```{r unrotated PCA}
# note we have NA values
# useful to figure out where they are in our modified dataset to figure out what to do with them

# first let's figure out how many
sum(is.na(q1data_mod)) #76 NAs
#that's a lot of missing values... are they full rows or individual observations?
which(is.na(q1data_mod))
#from cursory look, looks like the missing NAs are full rows
#I've elected to drop them
q1data_mod <- q1data_mod %>% 
  na.omit()

# now we can run PCA
pca <- princomp(q1data_mod, cor = TRUE)

# scree plot to visualize what's happening
plot(pca, type = "lines", main = "Scree Plot")

# extract eigenvalues:
eigens <- eigen(og_cor)
eigval <- eigens$values
eigvec <- eigens$vectors
```

At first glance, from the scree plot, it looks like we may have 5 components that explain most of the variance. However, only 4 of these have eigenvalues greater than 1.

The next step is to calculate our weighted component loadings.

```{r weighted loadings}
# Components 1-4 (eigval > 1)
pca$loadings[,1:4]

# Compute weighted loadings: 
# Weighted loadings = loadings * sqrt(eigenvalue)
loadings <- pca$loadings[,1:4]

# sqrt of the variance/eigenvalue
weights <- vec2diag(pca$sdev[1:4]) 
# remember: Multiplying by this matrix = multiplying each column of your loadings by sdev.

# component loadings 
loadings[, 1:4] %*% weights

loadings(principal(og_cor, nfactors = 19, rotate = "none"))
```

With the loadings calculated, we can reproduce our correlation matrix.

```{r rep cor mat}
# first multiply our eigenvectors and eigenvalues by t()
rep_cor <- round(eigvec[,1:4] %*% diag(eigens$values[1:4]) %*% t(eigvec[,1:4]), 2)
# extract residuals
# og cor - 4 best eigens
resid_cor <- round(og_cor - rep_cor, 2)
resid_cor
```

We now have our loadings from the first 4 vectors, corresponding to 4 components that explain the most variance (eigvalues > 1). Now we want to use this to consider our data, so we'll rotate using varimax orthogonal rotation.

```{r rotate orthog}
# specifying varimax
pca_var <- principal(q1data_mod, rotate = "varimax", nfactors = 4, missing = TRUE)
summary(pca_var)
pca_var$loadings
pca_var$residual

# calculating variance explained
round(eigens$values[1:4] / sum(eigens$values[1:4]), 4)
```

**How many components emerge, and how much variance do they explain all together?**
We had 4 components emerge, explaining 38.6%, 32.59%, 15.72%, and 13.09% of the variance respectively. If we add that up, our 4 components explain 99% of the variance.

**Are there any residual correlations (i.e., the difference between observed and reproduced) that are large (e.g., >0.2)?**

Yes. There are residual correlations greater than .2, all of which occur along the diagonal when the individual item is correlated with itself. However, this isn't meaningful to interpretation.

**Do they tend to cluster on one or the other component?**

There is one residual correlation of -.21 between items basf1 and basf3. When looking at the component loadings, these items mapped onto RC4 and RC3, which correspond to reward seeking and fun seeking, respectively.

**Based on the component loadings in the rotated solution, label each of the components.**  
* Component 1 (RC2) = Avoidant Behaviors (Avoidant + Opposite Fun/Drive)
* Component 2 (RC1) = Drive (Drive + Mostly not Avoidant + Some Reward/Fun)  
* Component 3 (RC4) = Reward Seeking (Reward + Some Drive)  
* Component 4 (RC3) = Fun Seeking (Fun + Some Drive + Opposite Avoidant)  


## Question 4

**Context:** A researcher hypothesizes that fluid intelligence is composed of two underlying factors:  
* Sequential processing  
* Simultaneous processing
She collected 3 measures of the first and 5 measures of the second, using 200 participants.

**Using R, run a basic CFA model whereby `sequential` is indicated by the first 3 items (`handmov`, `numbrec`, and `wordord`) and `simultaneous` is indicated by the last 5 items (`gesclos`, `triangle`, `spatmem`, `matanalg`, and `photser`).**

```{r CFA}
#establish our model
model <- '
SEQ  =~ handmov + numbrec + wordord
SIM =~ gesclos + triangle + spatmem + matanalg + photser

SEQ ~~ SIM
'
fit <- cfa(model, sample.cov = q4data, sample.nobs = 200)
summary(fit, fit.measures = TRUE)
```


**Report all the path weights (with SEs):**  
From `sequential`:  
* handmov 1.000  
* numbrec 1.147(0.181)  
* wordord 1.388(0.219)  

From `simultaneous`:  
* gesclos 1.000  
* triangle 1.445(0.227)  
* spatmem 2.029(0.335)  
* matanalg 1.212(0.212)  
* photser 1.727(0.265)

**Report the covariance between the factors:**  
1.271(0.324)

**Report (any) one fit index:**
Comparative Fit Index (CFI) = 0.959

*Note:* I tried to install the {semPlots} package, but it kept crashing RStudio for me. Once I got it to stop doing that, it kept prompting me to install {igraphs} which I did multiple times but still couldn't get semPlots to open. Are there any other alternatives you know of to plot these types of models using a different package and/or any tips you can give me if other students have run into this issue?
