---
title: "Problem Set 2"
author: "Janette Avelar"
date: "4/8/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The central limit theorem states that sample means will tend toward a normal distribution as sample sizes increase, regardless of the distribution from which the individual data points within each sample were drawn. Your homework assignment is to demonstrate this. Please create a script in R demonstrating the CLT by following the algorithm below.


### Step 1
Create a variable, “iter” indicating the number of times to iterate through the simulation (that is described in Step 3). Set that variable to be a large number (e.g., 10,000). Create another variable, “sample_size”, that contains a single, modest sample size (e.g., 50).

```{r set iter and sample size}
iter <- 10000
sample_size <- 50
```

### Step 2
Create an empty variable that will be used to store the mean of the sample that will be drawn in each iteration. The variable should be a vector with one row per sample mean that you will collect (i.e., one for each iteration through the loop).

```{r preallocation}
sample_means <- rep(NA_real_, iter)
```

### Step 3
Create a "for" loop that starts at 1 and goes through each sample iteration. Call the current iteration number `i`. Inside the loop:
a. Draw a sample of size "sample_size" from a random uniform distribution, and store this sample in a new variable called "sample".
b. Calculate the mean of the sample, and then store the mean in a variable called sample_mean
c. Store the sample mean in the *i*th place in the vector that you created in Step 2 (where *i* is the index number of the current sample iteration.)
d. End the loop.

```{r for loop}
for (i in 1:iter) {
  sample           <- runif(sample_size)
  sample_mean      <- mean(sample)
	sample_means[i]  <- sample_mean
}
```

### Step 4
After you've looped through all of the samples, create a histogram displaying the distribution of the sample means. Make sure the histogram has 100 bins.

```{r histogram}
hist(sample_means, 
     breaks = 100,
     main = paste("Sampling Distribution with Sample Size", sample_size), xlab = "Sample Means")
```

### Step 5
Calculate the standard deviation of the sample means using the `sd()` function, and then print it to the workspace to two decimal places using the `print()` function. In the printout, be sure to include the sample size of each individual sample.

```{r print SD}
sd <- sd(sample_means)
print(paste0("The standard deviation of the sample means is ", round(sd, 2), " for a sample size of ", sample_size, "."))
```

### BONUS
Embed the algorithm above in a second loop that repeats the entire procedure for 20 different sample sizes (of your choosing). That is, iterate steps 3-5. Plot the histograms in a 4x5 matrix of plots. Plot the standard deviations in a separate figure.

```{r nested for loop}
iter_b <- 1000
sample_size_b <- c(30, 60, 90, 120, 150, 180, 210, 240, 270, 300, 330, 360, 390, 420, 450, 480, 510, 540, 570, 600)

par(mfrow = c(4,5))

for(i in sample_size_b) {
  sample_means_b <- rep(NA_real_, iter_b)
  sample_sd_b    <- rep(NA_real_, iter_b)
  for(j in iter_b) {
    sample_b           <- runif(sample_size_b[i])
    sample_mean_b      <- mean(sample_b)
    sample_means_b[j]  <- sample_mean_b
    sample_sd_b[j]     <- sd(sample_means_b)
  }
  hist(sample_means_b, 
     main = paste("Sampling Distribution with Sample Size", sample_size_b[i]), 
     xlab = "Sample Means")
  hist(sample_sd_b,
       main = paste("Distribution of Standard Deviations with Sample Size", sample_size_b[i]),
       xlab = "Sample Standard Deviations")
}
```

*I gave it a shot but wasn't sure how to get it working.*

Modified nested loop - **ANSWER**:

```{r nested loop KEY}
iter_b <- 1000
sample_size_b <- c(30, 60, 90, 120, 150, 180, 210, 240, 270, 300, 330, 360, 390, 420, 450, 480, 510, 540, 570, 600)

par(mfrow = c(4,5))

for(i in 1:length(sample_size_b)) {
  sample_means_b <- rep(NA_real_, iter_b)
  sample_sd_b    <- rep(NA_real_, iter_b)
  for(j in 1:iter_b) {
    sample_b           <- runif(sample_size_b[i])
    sample_mean_b      <- mean(sample_b)
    sample_means_b[j]  <- sample_mean_b
    sample_sd_b[j]     <- sd(sample_means_b)
  }
  hist(sample_means_b, 
     main = paste("Sampling Distribution with Sample Size", sample_size_b[i]), 
     xlab = "Sample Means")
  hist(sample_sd_b,
       main = paste("Distribution of Standard Deviations with Sample Size", sample_size_b[i]),
       xlab = "Sample Standard Deviations")
}
```

