---
title: "Problem Set 1"
author: "Janette Avelar"
date: "4/2/2022"
output:
  word_document: default
  pdf_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(pracma)
library(MASS)
```

## Part 3: Checking Work Using R

**Woodward Appendix A.2.2**

1. Form the transpose of each matrix:

\[ a = \begin{bmatrix}
1 & 2 \\
3 & 4
\end{bmatrix} \]

```{r 1a transposed}
a_trans <- matrix(data = c(1, 2, 
                         3, 4), 
                nrow = 2, ncol = 2, byrow = TRUE) %>% 
  t() %>% 
  print()
```

\[ b = \begin{bmatrix}
1 \\
2 \\
3
\end{bmatrix} \]

```{r 1b transposed}
b_trans <- matrix(data = c(1,
                           2,
                           3),
                  nrow = 3, ncol = 1, byrow = TRUE) %>% 
  t() %>% 
  print()
```

\[ c = \begin{bmatrix}
1 & 4 & 3 \\
4 & 1 & 2 \\
3 & 2 & 1 \\
\end{bmatrix} \]

```{r 1c transposed}
c_trans <- matrix(data = c(1, 4, 3,
                            4, 1, 2,
                            3, 2, 1),
                   nrow = 3, ncol = 3, byrow = TRUE) %>% 
  t() %>% 
  print()
```

\[ d = \begin{bmatrix}
4 & 4 & 4 & 1
\end{bmatrix} \]

```{r 14 transposed}
d_trans <- matrix(data = c(4, 4, 4, 1),
                  nrow = 1, ncol = 4, byrow = TRUE) %>% 
  t() %>% 
  print()
```

2. Which of the above matrices are symmetric?

```{r a symmetry}
# testing a
dim(a_trans) == dim(t(a_trans))
a_trans == t(a_trans) #nope

# testing b
dim(b_trans) == dim(t(b_trans)) #nope

# testing c
dim(c_trans) == dim(t(c_trans))
c_trans == t(c_trans) #yes

# testing d
dim(d_trans) == dim(t(d_trans)) #nope
```

C and C' are symmetric.

**Woodward Appendix A.2.5**

1. Compute a'a and aa'.

```{r part 2 question 1}
# define a for part 2
a2 <- matrix(data = c(1,
                      2,
                      3),
             nrow = 3, ncol = 1, byrow = TRUE) %>% 
  print()

# define a transposed
a2_trans <- matrix(data = c(1,
                      2,
                      3),
             nrow = 3, ncol = 1, byrow = TRUE) %>% 
  t() %>% 
  print()

# multiply a'a
a2_trans %*% a2

# multiple aa'
a2 %*% a2_trans
```

2. Compute a'b and b'a.

```{r part 2 question 2}
# define b for part 2
b2 <- matrix(data = c(2, 1,
                      2, 3,
                      3, 4),
             nrow = 3, ncol = 2, byrow = TRUE) %>% 
  print()

# define b transposed
b2_trans <- matrix(data = c(2, 1,
                      2, 3,
                      3, 4),
             nrow = 3, ncol = 2, byrow = TRUE) %>% 
  t() %>% 
  print()

# multiply a'b
a2_trans %*% b2

# multiply b'a
b2_trans %*% a2
```

*Looks like I forgot to finish this question or somehow overwrote my work in the original document. Whoops.*

3. Compute cd and dc.

```{r part 2 question 3}
# define c for part 2
c2 <- matrix(data = c(1, 2, 3,
                      4, 5, 6,
                      7, 8, 9),
             nrow = 3, ncol = 3, byrow = TRUE) %>% 
  print()

# define d for part 2
d2 <- matrix(data = c(9, 8, 7,
                      6, 5, 4,
                      3, 2, 1),
             nrow = 3, ncol = 3, byrow = TRUE) %>% 
  print()

# multiply cd
c2 %*% d2

# multiply dc
d2 %*% c2
```

*Some multiplication/addition errors occurred, but on the right track by hand.*

4. Compute dI_3 and I_3d.

```{r part 2 question 4}
# define I3
I3 <- matrix(data = c(1, 0, 0,
                      0, 1, 0,
                      0, 0, 1),
             nrow = 3, ncol = 3, byrow = TRUE) %>% 
  print()

# multiply dI_3
d2 %*% I3

# multiply I_3d
I3 %*% d2
```

5. Compute e(c+d).

```{r part 2 question 5}
# define e for part 2
e2 <- matrix(data = c(1, 4, 1),
                  nrow = 1, ncol = 3, byrow = TRUE) %>% 
  print()

# compute e(c+d)
e2 %*% (c2 + d2)
```

6. Compute e(c-d).

```{r part 2 question 6}
e2 %*% (c2 - d2)
```

*This one was off when conducted by hand.*

7. Compute ec-ed.

```{r part 2 question 7}
(e2 %*% c2) - (e2 %*% d2)
```

*This one was off when conducted by hand.*

8. Compute cdb.

```{r part 2 question 8}
c2 %*% d2 %*% b2
```

*This one was off when conducted by hand.*

**Woodward Appendix A.2.9**

1. Verify that aa^-1 = I_2 in the preceding example.

```{r part 3 question 1}
# define a for part 3
a3 <- matrix(data = c(3, 2,
                     2, 4),
            nrow = 2, ncol = 2, byrow = TRUE) %>% 
  print()

# define a inverse
a3_inv <- matrix(data = c(0.5, -0.25,
                          -0.25, 0.375),
                 nrow = 2, ncol = 2, byrow = TRUE) %>% 
  print()

# define I_2
I2 <- matrix(data = c(1, 0,
                      0, 1),
             nrow = 2, ncol = 2, byrow = TRUE) %>% 
  print()

# compute aa^-1
a3 %*% a3_inv

# compare to I_2
dim(a3 %*% a3_inv) == dim(I2)
(a3 %*% a3_inv) == I2
```

2. Compute the inverse of b, where 
\[ b=\begin{bmatrix}
6 & 1 \\
1 & 3
\end{bmatrix} \]

```{r part 3 question 2}
# define b for part 3
b3 <- matrix(data = c(6, 1,
                      1, 3),
             nrow = 2, ncol = 2, byrow = TRUE) %>% 
  print()

# compute the inverse
solve(b3)

# convert to fraction to see if it matched
fractions(solve(b3))
```

3. Verify that bb^-1 = I_2 in part (b).

```{r part 3 question 3}
# assign b inverse
b3_inv <- solve(b3) %>% 
  print()

# compute bb^-1
bb1 <- round(b3 %*% b3_inv) %>% 
  print()

# compare to I_2
dim(b3 %*% b3_inv) == dim(I2)
bb1 == I2
```

**Reduced Row Echelon Form**

Determine the reduced row echelon form of the matrix:

\[ \begin{bmatrix}
2 & 1 & 2 \\
1 & -1 & 2 \\
-1 & -5 & 2 \\
-1 & -5 & 2 \\
-1 & 10 & -8
\end{bmatrix} \]

```{r reduced row echelon form}
# define the matrix
mat <- matrix(data = c(2, 1, 2,
                       1, -1, 2,
                       -1, -5, 2,
                       -1, -5, 2,
                       -1, 10, -8),
              nrow = 5, ncol = 3, byrow = TRUE) %>% 
  print()

# convert to rref
rref(mat)

# compare to answer
fractions(rref(mat))
```

*I was way off, oh well.*